The Host Galaxies of Rapidly Evolving Transients in the Dark Energy Survey

### 2020-03-04 RET_Hosts.tex

% Abstract of the paper
\begin{abstract}
Rapidly evolving transients (RETs) are a mysterious class of astrophysical event. They are characterised by lightcurves that decline much faster than standard supernovae (SNe), span vast ranges in peak luminosity and can be seen to redshifts greater than 1. Their evolution on fast timescales has hindered high quality follow-up observations, such that their origin and explosion/emission mechanism remains unexplained. In this paper, we investigate the host galaxies of the largest RET sample to date from the Dark Energy Survey (DES). Using deep-stacked photometry and emission-lines from OzDES spectroscopy, we derive host galaxy stellar mass, star-formation rate (SFR), and metallicity for 42 hosts. We find that RETs explode almost exclusively in star-forming galaxies and are thus likely associated with massive stars. Comparing RET hosts to samples of host galaxies of other explosive transient as well as field galaxies, we find that RETs prefer galaxies with high specific SFR, indicating a link to young stellar populations stripped-envelope SNe. RET hosts appear to show a lack of chemical enrichment, their metallicities akin to long duration gamma-ray bursts and superluminous SNe host galaxies. There are no clear relationships between properties of the host galaxies and the peak magnitudes or decline rates of the transients themselves.

@@

\subsubsection{RETs \label{subsubsec:compare_rets}}
Since the DES sample of RETs is by far the largest discovered to date, there is no other large sample of RETs with which to compare host galaxy properties. \citet{Drout2014} present host galaxies of 10 RETs discovered in the Pan-STARRS survey, with measurements of stellar masses and SFRs. To this we add the low-redshift transient AT2018cow with host galaxy photometric measurements from \citet{Perley2019} and metallicity (\citealt{Pettini2004} O3N2) from \citet{Morokuma-Matsui2019}, and SN2018gep with data from \citet{Ho2019}.

@@

In order to subtract the stellar component of the host galaxy spectra, we use the Penalized PiXel-Fitting software (\texttt{pPXF}; \citealt{Cappellari2004,Cappellari2012,Cappellari2017}), using the MILES library of single stellar populations \citep{Vazdekis2010}. By subtracting the best-fitting composite stellar spectrum from the \texttt{pPXF} fit, we are left with a `gas' spectrum, comprising the emission lines. An example of this procedure is shown in Fig. \ref{fig:host_spec}. We fit the emission lines using \texttt{PySpecKit} \citep{Ginsburg2011}. In order to estimate the uncertainty on the emission line fluxes, we fit $10^4$ realisations of the line, each time adding perturbations to the line by drawing from a Gaussian distribution centered on 0, and with a standard deviation equal to the root-mean-square (RMS) of the measured flux noise in a 100~\AA~window around the central wavelength of the line. We take the mean and standard deviation of the resulting fits as our flux and its uncertainty, respectively.

@@

Other local rapid transients include SN2018gep \citep{Ho2019}, a spectroscopically classified SN Ic-bl with a rapid rise. The host of SN2018gep appears more similar to the DES RET sample, lying in the same $M_*$-SFR and $M*$-sSFR plane, as well as lying below the MZR. While the SN2018gep host is lower in stellar mass than any DES RET ($\log \left(M/M_{\odot}\right) =8.11$, galaxies of that mass are unlikely to have been detected at the redshifts of the DES RETs \citep{Wiseman2020}. The authors' conclusion that SN2018gep is related to a shock-breakout of a massive, stripped-envelope star is similar to that posited in Section \ref{subsec:disc_origin}. SN2018kzr \citep{McBrien2019} is one of the most rapidly declining transients ever discovered, with spectral signatures similar to SNe Ic. While host galaxy properties are not derived, the authors of that paper refer to narrow emission from the host galaxy, along with an apparently small, blue, star-forming host.

@@

This work makes extensive use of Astropy,\footnote{http://www.astropy.org} a community-developed core Python package for Astronomy \citep{AstropyCollabor

>>>>

% Abstract of the paper
\begin{abstract}
Rapidly evolving transients (RETs), often also termed fast blue optical transients (FBOTs), are a mysterious class of astrophysical event. They are characterised by lightcurves that decline much faster than standard supernovae (SNe), span vast ranges in peak luminosity and can be seen to redshifts greater than 1. Their evolution on fast timescales has hindered high quality follow-up observations, such that their origin and explosion/emission mechanism remains unexplained. In this paper, we investigate the host galaxies of the largest RET sample to date from the Dark Energy Survey (DES). Using deep-stacked photometry and emission-lines from OzDES spectroscopy, we derive host galaxy stellar mass star-formation rate (SFR) for 49 hosts, and metallicity for 42. We find that RETs explode almost exclusively in star-forming galaxies and are thus likely associated with massive stars. Comparing RET hosts to samples of host galaxies of other explosive transient as well as field galaxies, we find that RETs prefer galaxies with high specific SFR, indicating a link to young stellar populations, similar to stripped-envelope SNe. RET hosts appear to show a lack of chemical enrichment, their metallicities akin to long duration gamma-ray bursts and superluminous SNe host galaxies. There are no clear relationships between properties of the host galaxies and the peak magnitudes or decline rates of the transients themselves.

@@

\subsubsection{RETs \label{subsubsec:compare_rets}}
Since the DES sample of RETs is by far the largest discovered to date, there is no other large sample of RETs with which to compare host galaxy properties. \citet{Drout2014} present host galaxies of 10 RETs discovered in the Pan-STARRS survey, with measurements of stellar masses and SFRs. To this we add the low-redshift transient AT2018cow (nicknamed ``The Cow'') with host galaxy photometric measurements from \citet{Perley2019} and metallicity (\citealt{Pettini2004} O3N2) from \citet{Morokuma-Matsui2019}, SN2018gep with data from  \citet{Ho2019}, and ZTF18abvkwla (nicknamed ``The Koala'') from \citet{Ho2020}.

@@

In order to subtract the stellar component of the host galaxy spectra, we use the Penalized PiXel-Fitting software (\texttt{pPXF}; \citealt{Cappellari2004,Cappellari2012,Cappellari2017}), using the MILES library of single stellar populations \citep{Vazdekis2010}. By subtracting the best-fitting composite stellar spectrum from the \texttt{pPXF} fit, we are left with a `gas' spectrum, comprising the emission lines. An example of this procedure is shown in Fig. \ref{fig:host_spec}. We fit the emission lines with Gaussian profiles. In order to estimate the uncertainty on the emission line fluxes, we fit $10^4$ realisations of the line, each time adding perturbations to the line by drawing from a Gaussian distribution centered on 0, and with a standard deviation equal to the root-mean-square (RMS) of the measured flux noise in a 100~\AA~window around the central wavelength of the line. We take the mean and standard deviation of the resulting fits as our flux and its uncertainty, respectively.

@@

Other local rapid transients include SN2018gep \citep{Ho2019}, a spectroscopically classified SN Ic-bl with a rapid rise. The host of SN2018gep appears more similar to the DES RET sample, lying in the same $M_*$-SFR and $M*$-sSFR plane, as well as lying below the MZR. While the SN2018gep host is lower in stellar mass than any DES RET ($\log \left(M/M_{\odot}\right) =8.11$, galaxies of that mass are unlikely to have been detected at the redshifts of the DES RETs \citep{Wiseman2020}. The authors' conclusion that SN2018gep is related to a shock-breakout of a massive, stripped-envelope star is similar to that posited in Section \ref{subsec:disc_origin}.

The rapidly evolving lightcurve of ZTF18abvkwla (``the Koala'') has been attributed to shock interaction, while radio emission can be explained by a collimated jet. The host of the Koala is a low metallicity starburst more typical of LGRBs and SLSNe, and places this transient at the very extreme end of the DES RET host population. While we note that the \citet{Ho2020} study made multiple non-detections of radio emission from the DES RETs, these were taken at very late epochs ($\geq 1\mathrm{year}$), so the presence of jets in the early evolution is not ruled out. Seen from the other point of view, we cannot rule out that the Koala comes from the same population of transients as the DES RETs.

SN2018kzr \citep{McBrien2019} is one of the most rapidly declining transients ever discovered, with spectral signatures similar to SNe Ic. While host galaxy properties are not derived, the authors of that paper refer to narrow emission from the host galaxy, along with an apparently small, blue, star-forming host.

@@

This work makes extensive use of Astropy,\footnote{http://www.astropy.org} a community-developed core Python package for Astronomy \citep{AstropyCollaboration2013,AstropyCollaboration2018}, Pandas \citep{Mckinney2010}, and matplotlib \citep{Hunter2007}.
